Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-22 submission/manuscript.Rmd
add second result section update methods

@@

**Random Forest models with SCFA concentrations do not classify tumor better.**

@@

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. We next assessed whether SCFAs added to Random Forest models built with OTU data significantly improved classification. Next, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002]. Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm. The measured SCFA concentrations were first tested for differences between groups. The ability of 16S rRNA gene sequencing to classify these concentrations were then assessed. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995.

>>>>

# Load in data for select genes in SCFA pathways
picrust_select <- read_csv("../data/process/specific_scfa_kruskal_picrust_summary.csv")
opf_select <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv")

@@

**Random Forest models with SCFA concentrations do not classify tumor better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations. For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although there was no added information gained for classification of disease by including SCFAs to Random Forest models, it is possible that the genes for enzymes involved in SCFA pathways may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.

@@

**Random Forest Models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for both classification and regression models [@randomforest_citation_2002]. This was repeated on 100 differen 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. Classification models were built to group normal versus adenoma and normal versus carcinoma or high versus low SCFA concentrations. Regression models were built to assess how well OTUs could be used to get the SCFA concentration within a sample.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest. Regression models also were created to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm. This was done to assess the ability of 16S rRNA gene sequencing to classify the SCFA concentrations.

### 2018-05-22 submission/manuscript.tex
add second result section update methods

\textbf{Decreased SCFA concentrations are not associated with tumors.}
There was no difference between individuals with normal colons and
patients with either an adenoma or carcinoma after multiple comparison
correction for any of the SCFAs measured (P-value \textgreater{} 0.15)
{[}Figure 1A - 1C{]}. Although there was a trend for increasing acetate,
butyrate, and propionate concentrations after treatment for tumors,
there was no significant difference pre- and post-treatment for
individuals with adenoma (P-value \textgreater{} 0.20) or carcinoma
(P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (15). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. We next assessed whether SCFAs added to Random
Forest models built with OTU data significantly improved classification.
Next, models to classify high or low SCFA concentration based on 16S
rRNA gene sequencing data were created using Random Forest (16).
Regression models to classify the exact SCFA concentration based on 16S
rRNA gene sequencing data also were built using the Random Forest
algorithm. The measured SCFA concentrations were first tested for
differences between groups. The ability of 16S rRNA gene sequencing to
classify these concentrations were then assessed. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Where appropriate Benjamini-Hochberg was
used to correct for multiple comparisons {[}(17).

@@

\hypertarget{ref-r_citation_2017}{}
15. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-randomforest_citation_2002}{}
16. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

>>>>

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used frozen fecal samples from 490 individuals and HPLC to measure
acetate, butyrate, and propionate concentrations at a cross-sectional
point in time. There was no difference between individuals with normal
colons and patients with either an adenoma or carcinoma for any of the
SCFAs measured after multiple comparison correction (P-value
\textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next measured the
concentration of SCFAs in 67 patients with an adenoma or carcinoma in
which we had pre- and post-treatment fecal samples. Although there was a
general trend for increasing acetate, butyrate, and propionate
concentrations after treatment for tumors, there was no significant
difference pre- and post-treatment for patients with adenoma (P-value
\textgreater{} 0.20) or carcinoma (P-value \textgreater{} 0.80)
{[}Figure 1D - 1F{]}. Even though there was no change in SCFA
concentrations between individuals with normal colons and those with
tumors, this information could still be important to help classify
disease.

\textbf{Random Forest models with SCFA concentrations do not classify
tumor better.} Using the Random Forest algorithm we built models to
classify normal versus adenoma and normal versus carcinoma with OTUs or
OTUs and SCFA concentrations. For both adenoma and carcinoma models,
there was no difference between the median AUC of models with or without
SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
Although there was no added information gained for classification of
disease by including SCFAs to Random Forest models, it is possible that
the genes for enzymes involved in SCFA pathways may change due to
disease status; where a smaller number of microbes are responsible for
the observed SCFA concentrations.

@@

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(15). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations. Regression models were built to
assess how well OTUs could be used to get the SCFA concentration within
a sample.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (16). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (17). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using Random Forest. Regression models also were created to classify the
exact SCFA concentration based on 16S rRNA gene sequencing data also
were built using the Random Forest algorithm. This was done to assess
the ability of 16S rRNA gene sequencing to classify the SCFA
concentrations.

@@

\hypertarget{ref-randomforest_citation_2002}{}
15. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

\hypertarget{ref-r_citation_2017}{}
16. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.
